---
title: "Survey Q-Sort Estimator Recovery: Simulation Results"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dpi = 150, fig.path = "docs/figures/results/", fig.width = 8, fig.height = 4.5
)
source("R/qsort.R")
library(ggplot2)
library(scales)

study   <- readRDS("results/study_results.rds")
scored  <- study$scored
design  <- study$design
levels  <- study$levels
n_reps  <- study$replications

# A consistent display order and labelling for the estimators.
estimator_levels <- c("naive_mean", "naive_steep", "naive_boot",
                      "dirichlet", "pl_ranking", "pl_trials")
estimator_labels <- c(
  naive_mean  = "Naive mean (5-4-3-2-1)",
  naive_steep = "Naive mean (steep codes)",
  naive_boot  = "Naive mean (bootstrap SE)",
  dirichlet   = "Dirichlet-multinomial",
  pl_ranking  = "PL single ranking",
  pl_trials   = "PL four trials"
)
relabel <- function(x) factor(estimator_labels[x], levels = estimator_labels[estimator_levels])

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  legend.position = "bottom",
                  plot.title.position = "plot"))
pal <- c(
  "Naive mean (5-4-3-2-1)"   = "#B23A48",
  "Naive mean (steep codes)" = "#E08A3C",
  "Naive mean (bootstrap SE)"= "#7E5A9B",
  "Dirichlet-multinomial"    = "#2B7A78",
  "PL single ranking"        = "#3A6EA5",
  "PL four trials"           = "#1B998B"
)
```

This report is generated from a real run of `run_study.R`:
**`r nrow(design)` conditions × `r n_reps` replications = `r format(nrow(design) * n_reps, big.mark = ",")` simulated samples**,
each scored against a known planted truth. Every estimator returns a share for
each item plus a standard error, and confidence intervals are built as
`estimate ± z · se`, so coverage can be checked at any nominal level.

## The question

The survey Q-sort scores items by averaging `5-4-3-2-1` band codes. That assumes
the bands are evenly spaced and that every item belongs on the forced shape.
Real importance is lumpy, so the worry is that the average **compresses the
spacing** — shrinking the leaders, inflating the tail — and reports tight
intervals centred on the wrong number. We ask, across realistic conditions:
**which estimator recovers the true spacing, and whose intervals are honest?**

## Design grid

```{r design-grid}
design %>%
  transmute(
    `n items` = n_items, `n respondents` = n_resp,
    lumpiness, noise
  ) %>%
  distinct() %>%
  knitr::kable()
```

`lumpiness` controls how concentrated the truth is (higher = one item runs away
with it); `noise` is a temperature on respondents' choices (higher = noisier).
`pl_ranking` is attempted only where the Q-sort middle band stays a small tie
(see [Scalability](#scalability-why-pl_ranking-only-appears-for-short-lists)).

## Headline: spacing recovery

The clearest symptom is what each estimator does to the **leader** — the single
most important item. Pooled across the grid, here is the average recovered
leader share against the truth.

```{r leader-recovery}
leader <- scored %>%
  group_by(condition) %>%
  filter(true_share == max(true_share)) %>%
  ungroup() %>%
  group_by(estimator) %>%
  summarise(
    `recovered leader share` = mean(estimate),
    `true leader share`      = mean(true_share),
    `ratio (recovered/true)` = mean(estimate) / mean(true_share),
    .groups = "drop"
  ) %>%
  mutate(estimator = factor(estimator, estimator_levels)) %>%
  arrange(estimator) %>%
  mutate(estimator = estimator_labels[as.character(estimator)])
knitr::kable(leader, digits = 3)
```

A ratio below 1 is compression (the leader is understated); above 1 is
over-shooting. The naive family and the Dirichlet sit well below 1 — they
**flatten the leader**. The single-ranking Plackett-Luce swings the other way
and **over-states** it. The four-trial Plackett-Luce lands closest to 1.

```{r bias-by-stratum, fig.height=4}
bias_tab <- summarise_recovery(scored) %>%
  mutate(estimator = relabel(estimator),
         stratum = factor(stratum, c("extreme", "middle")))

ggplot(bias_tab, aes(estimator, bias, fill = estimator)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  facet_wrap(~ stratum, labeller = as_labeller(c(extreme = "Extreme items (top/bottom)",
                                                 middle = "Middle items"))) +
  scale_fill_manual(values = pal, guide = "none") +
  coord_flip() +
  labs(title = "Bias in recovered share, by estimator and item stratum",
       subtitle = "Negative = compresses the item's share; positive = inflates it",
       x = NULL, y = "Mean (estimate - truth)")
```

For the **extreme** items — the ones decisions hinge on — the naive estimators
carry a clear negative bias (they shrink the top and bottom toward the middle),
while the single-ranking PL carries a positive bias (it inflates the leader).
The four-trial PL sits nearest zero in both strata.

## Honesty of the intervals

A 95% interval should contain the truth 95% of the time. Plotting empirical
coverage against the nominal level, an honest method tracks the diagonal; an
overconfident one falls below it.

```{r coverage-calibration, fig.height=4.6}
cov <- summarise_coverage(scored, levels) %>%
  mutate(estimator = relabel(estimator),
         stratum = factor(stratum, c("extreme", "middle")))

ggplot(cov, aes(nominal, coverage, colour = estimator)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ stratum, labeller = as_labeller(c(extreme = "Extreme items (top/bottom)",
                                                 middle = "Middle items"))) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0.4, 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(title = "Interval calibration: empirical coverage vs nominal level",
       subtitle = "On the dashed line = honest; below it = overconfident",
       x = "Nominal level", y = "Empirical coverage") +
  guides(colour = guide_legend(nrow = 2))
```

```{r coverage-95-table}
summarise_coverage(scored, 0.95) %>%
  select(-nominal) %>%
  pivot_wider(names_from = stratum, values_from = coverage) %>%
  mutate(estimator = factor(estimator, estimator_levels)) %>%
  arrange(estimator) %>%
  transmute(Estimator = estimator_labels[as.character(estimator)],
            `Extreme coverage` = extreme, `Middle coverage` = middle) %>%
  knitr::kable(digits = 3, caption = "Empirical coverage of nominal 95% intervals.")
```

The naive family and the Dirichlet under-cover badly on the extreme items: their
intervals are reasonably sized but centred on the compressed estimate, so they
miss the truth far more often than advertised. Critically, the **bootstrap
version is no better than the analytic one** — ruling out "we just sized the
intervals wrong." The bias is the culprit. The single-ranking PL also
under-covers (it over-shoots instead). The four-trial PL is the only estimator
that approaches its advertised rate.

## Accuracy (RMSE)

```{r rmse-table}
summarise_recovery(scored) %>%
  select(estimator, stratum, rmse, width95) %>%
  pivot_wider(names_from = stratum, values_from = c(rmse, width95)) %>%
  mutate(estimator = factor(estimator, estimator_levels)) %>%
  arrange(estimator) %>%
  transmute(
    Estimator = estimator_labels[as.character(estimator)],
    `RMSE extreme` = rmse_extreme, `RMSE middle` = rmse_middle,
    `95% width extreme` = width95_extreme, `95% width middle` = width95_middle
  ) %>%
  knitr::kable(digits = 4)
```

The four-trial PL has the lowest error on the extreme items by a wide margin.
The naive estimators look precise (narrow widths) but that precision is around
the wrong value — exactly the trap the study set out to expose.

## How the conclusion holds across conditions

Extreme-item coverage of nominal 95% intervals, broken out by the design
factors, confirms the pattern is not an artefact of one cell.

```{r coverage-by-condition, fig.height=4.6}
by_cond <- scored %>%
  filter(stratum == "extreme") %>%
  group_by(estimator, n_items, n_resp, lumpiness, noise) %>%
  summarise(coverage = mean(abs(error) <= z_for(0.95) * se), .groups = "drop") %>%
  mutate(estimator = relabel(estimator),
         panel = paste0("n_items=", n_items, ", N=", n_resp))

ggplot(by_cond, aes(factor(lumpiness), coverage, colour = estimator, shape = factor(noise))) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.4, position = position_dodge(width = 0.4)) +
  facet_wrap(~ panel) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_shape_manual(values = c(`1` = 16, `2` = 1), name = "noise") +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(title = "Extreme-item coverage of 95% intervals, by condition",
       subtitle = "Dashed line = the advertised 95%",
       x = "lumpiness") +
  guides(colour = guide_legend(nrow = 2))
```

## Scalability: why `pl_ranking` only appears for short lists

The single-ranking collapse has to encode the Q-sort's middle band as a single
**tie of order `n_items − 6`**, and PlackettLuce's likelihood cost grows steeply
with that order. The four-trial factorization never forms that tie. A quick
wall-clock benchmark of a single `pl_ranking` fit makes the wall obvious:

```{r pl-scaling, fig.height=3.6}
bench <- tibble(
  n_items = c(10, 12, 14, 16),
  tie_order = n_items - 6,
  seconds = c(NA, NA, NA, NA)
)
# Measure the short, fast cases live; cite the slow ones (timings from this
# environment) so the report knits quickly.
measured <- map_dbl(c(10, 12), function(ni) {
  set.seed(1)
  tr <- make_truth(ni, 1.4); q <- simulate_qsort(tr, 300, 1)
  t0 <- Sys.time(); invisible(fit_pl_ranking(q)); as.numeric(Sys.time() - t0, units = "secs")
})
bench$seconds[bench$n_items == 10] <- measured[1]
bench$seconds[bench$n_items == 12] <- measured[2]
bench$seconds[bench$n_items == 14] <- 53   # measured separately; ~53 s
bench$seconds[bench$n_items == 16] <- 75   # exceeded a 75 s cap (lower bound)

ggplot(bench, aes(tie_order, seconds)) +
  geom_line(colour = "#3A6EA5") +
  geom_point(size = 3, colour = "#3A6EA5") +
  geom_text(aes(label = paste0("n=", n_items)), vjust = -0.8, size = 3.4) +
  scale_y_continuous(limits = c(0, 85)) +
  labs(title = "pl_ranking fit time explodes with the middle-band tie order",
       subtitle = "Tie order = n_items - 6; n=16 exceeded a 75 s cap",
       x = "Middle-band tie order", y = "Seconds for one fit")
```

So even setting aside which likelihood is more faithful, the single-ranking
model is **computationally infeasible for realistic list lengths** (it did not
finish for `n_items ≥ 16` in our environment), whereas the four-trial model fits
in a fraction of a second throughout.

## Conclusion

```{r conclusion-numbers}
lead_ratio <- leader$`ratio (recovered/true)`
names(lead_ratio) <- leader$estimator
cov95 <- summarise_coverage(scored, 0.95) %>% filter(stratum == "extreme")
cov_pl <- round(100 * cov95$coverage[cov95$estimator == "pl_trials"])
cov_naive <- round(100 * cov95$coverage[cov95$estimator == "naive_mean"])
```

Across `r format(nrow(design) * n_reps, big.mark = ",")` simulated samples:

- **The naive mean compresses the spacing.** It recovers the leader at about
  `r round(100 * lead_ratio["Naive mean (5-4-3-2-1)"])`% of its true share, and
  its nominal-95% intervals cover the extreme items only about
  `r cov_naive`% of the time. Swapping in steeper codes shifts the answer,
  confirming the equal-spacing assumption is doing real work; the cluster
  bootstrap does not rescue coverage, confirming the fault is the estimate, not
  the interval.
- **The single-ranking Plackett-Luce over-corrects** — it inflates the leader
  and is also miscalibrated — and it does not scale to realistic list lengths.
- **The four-trial Plackett-Luce is the estimator to use.** It recovers the
  leader at about `r round(100 * lead_ratio["PL four trials"])`% of truth, has
  the lowest RMSE on the items that matter, comes closest to honest coverage
  (about `r cov_pl`% on the extreme items at the nominal 95%), and fits quickly
  at every list length.

The four-trial model is not perfect — its extreme-item coverage still sits below
the advertised rate under the noisiest conditions — but it is the only candidate
that recovers the spacing without either flattening or inflating it, and the only
principled one that runs at scale.
